**Measuring Depression and Anxiety with 4 items? Adaptation of the PHQ-4 to increase its Sensitivity to Subclinical Variability**

### 2025-06-06 paper/5_Collabra/manuscript.qmd
clarified the use of the trait subscale of the STAI-5 + IAS acronym

This brief report aims at testing the possibility of enhancing - with minimal changes to the original scale - the PHQ-4 sensitivity to mild mood level inflections. In the first study, we will evaluate whether the new response option is prevalently used by participants, and whether it does capture a specific part of the construct. In the second study, we will compare the refined PHQ-4 version to the original one in terms of sensitivity to subclinical variability in depression, using the Beck Depression Inventory [BDI-II, @beck1996beck] and the State Trait Anxiety Inventory [STAI-5, @zsido2020development] as our ground-truth measures of depression and anxiety.

@@

The initial sample consisted of 1053 participants, recruited (181 were recruited on *Prolific*, 772 students from the University of Sussex via *SONA*, and the rest through convenience sampling as part of dissertation students' data collection).
We used attention checks as the primary target for participant exclusion. 
We excluded 194 participants (18.42\%) for failing at least one attention check, and 23 (2.18\%) that were outliers ($|z_{robust}| > 2.58$) on measures significantly related to the probability of failing attention checks [namely, the standard deviation of all the items of the IAS, as well as the the multivariate distance obtained with the OPTICS algorithm, see @theriault2024check].
The experiment duration was not related to the probability of failing attention checks and was thus not used as an exclusion criterion.

@@

The short version of the State-Trait Anxiety Inventory [STAI-5, @zsido2020development] was used as a ground truth measure of anxiety. This abridged version of the STAI [@spielberger1970manual] includes 5 items rated on a 4-point Likert scale. Changes were made in the instructions from asking *"how participants feel right now"* to *"over the past 2 weeks"* to keep it consistent with the instructions of the PHQ-4 and BDI-II. A general score of anxiety was computed by averaging all the items.

Participants were also asked to complete two questionnaires of interoception, namely the Interoceptive Accuracy Scale [IAS - 21 items rated on analog scales, @murphy2020testing] and the Multidimensional Assessment of Interoceptive Awareness [MAIA-2 - 37 items, @mehling2018multidimensional].

@@

The original or refined version of the PHQ-4 was followed by the BDI-II, STAI-5, IAS, and MAIA-2, presented in random order. The IAS [@murphy2020testing] and the MAIA-2 [@mehling2018multidimensional] were included as part of another focused on interoception, and were only used in this study as part of data quality control checks.

@@

```{r}
#| label: fig-two
#| fig-cap: "A) PHQ-4 depression and anxiety scores against their respective ground-truth measures, the BDI-22 and the STAI-5. Bayes factors in grey tell if there is a difference, for the same PHQ-4 score, between the original and the refined version (BFs < 1 suggest no difference and thus evidence for a comparability of the refined version with respect fo the original scale. Bayes factors in yellow represent how new in-between scores (0.5, 1.5, 2.5, ...) available with refined version differ from the adjacent scores (BFs > 3 suggest that half a point of difference on the refined PHQ-4 relates to a significant difference on the ground truth measure). BF < 1/3°, BF > 3\\*, BF > 10\\*\\*, BF > 30\\*\\*\\*. B) Bootstrapped distributions of the difference of correlation between the revised PHQ-4 scores and the original one for sub-clinical threshold scores of depression and anxiety. Positive differences suggest that the correlation between the ground-truth measure and the refined PHQ-4 score was stronger compared to the original version. C) Predictive power of the PHQ-4 scores on the presence of a depression or anxiety disorder. The upper plots show the relationship modelled by a logistic regression, while the above plots represent the ROC curves (in which a line further away from the diagonal represents a higher combination of sensitivity and specificity)."
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"

@@

### PHQ-4 Anxiety *vs.* STAI-5

The linear regression predicting the STAI-5 general score with the PHQ-4 anxiety score showed no interaction related to the PHQ-4 version ($\Delta Intercept_{\text{refined}} =  -0.02,~95\%~CI~[-0.15, 0.11],~t(832) =  -0.32,~p = 0.750$; $\Delta \beta_{\text{refined}} = 0.01,~95\%~CI~[-0.03, 0.05],~t(832) = 0.56,~p = 0.576$), suggesting no differences in the relationship pattern between the two versions.

Moreover, Bayesian *t*-tests comparing the STAI-5 scores between the refined and the original version at each integer score yielded no evidence in favour of a significant difference. In other words, having the same score on the refined version as on the original version was related to the same outcome on the STAI-5.

However, comparing in-between scores with adjacent scores yielded mixed results. Scoring 0.5 on the PHQ-4 anxiety was not significantly associated with a different level of STAI-5 compared to scoring 0 (BF = 1.83), but was with scores of 1 (BF > 30). Similarly, there was no evidence that scoring 1.5 was different from scoring 1 (BF = 0.605), but strong evidence that it was different from scoring 2 (BF > 30).

@@

While the relationship pattern (i.e., the slope of the linear relationship) was not affected by the PHQ-4 version, we focused next on testing the difference in the strength (i.e., the precision) of the relationship, in particular at the lower end of the spectrum (i.e., for sub-clinical threshold scores of the BDI-II and STAI-5). We bootstrapped (2000 iterations) the difference in correlation between the refined and the original version for each of the two ground-truth measures, separately for the BDI-II subsamples (minimal to mild <= 18; moderate to severe > 18) and the STAI-5 subsamples (minimal to mild < 2; moderate to severe >= 2).

@@

For the STAI-5, there was no difference in the correlation between the refined and the original version in the subclinical range of the STAI-5. Surprisingly, we observed a stronger correlation between the refined PHQ-4 Anxiety score and the STAI-5 in the moderate to severe range compared to the original version ($p_{one-sided} = 0.017$).

>>>>

This brief report aims at testing the possibility of enhancing - with minimal changes to the original scale - the PHQ-4 sensitivity to mild mood level inflections. In the first study, we will evaluate whether the new response option is prevalently used by participants, and whether it does capture a specific part of the construct. In the second study, we will compare the refined PHQ-4 version to the original one in terms of sensitivity to subclinical variability in depression, using the Beck Depression Inventory [BDI-II, @beck1996beck] and the Trait scale of the State Trait Anxiety Inventory [STAIT-5, @zsido2020development] as our ground-truth measures of depression and anxiety.

@@

The initial sample consisted of 1053 participants, recruited (181 were recruited on *Prolific*, 772 students from the University of Sussex via *SONA*, and the rest through convenience sampling as part of dissertation students' data collection).
We used attention checks as the primary target for participant exclusion. 
We excluded 194 participants (18.42\%) for failing at least one attention check, and 23 (2.18\%) that were outliers ($|z_{robust}| > 2.58$) on measures significantly related to the probability of failing attention checks [namely, the standard deviation of all the items of the Interoceptive Accuracy Scale, as well as the the multivariate distance obtained with the OPTICS algorithm, see @theriault2024check].
The experiment duration was not related to the probability of failing attention checks and was thus not used as an exclusion criterion.

@@

The short version of the trait scale of the State-Trait Anxiety Inventory [STAIT-5, @zsido2020development] was used as a ground truth measure of anxiety. This abridged version of the STAI [@spielberger1970manual] includes 5 items rated on a 4-point Likert scale. Changes were made in the instructions from asking *"how participants feel right now"* to *"over the past 2 weeks"* to keep it consistent with the instructions of the PHQ-4 and BDI-II. A general score of anxiety was computed by averaging all the items.

Participants were also asked to complete two questionnaires of interoception, namely the Interoceptive Accuracy Scale [IAS, 21 items rated on analog scales, @murphy2020testing] and the Multidimensional Assessment of Interoceptive Awareness [MAIA-2 - 37 items, @mehling2018multidimensional].

@@

The original or refined version of the PHQ-4 was followed by the BDI-II, STAIT-5, IAS, and MAIA-2, presented in random order. The IAS [@murphy2020testing] and the MAIA-2 [@mehling2018multidimensional] were included as part of another focused on interoception, and were only used in this study as part of data quality control checks.

@@

```{r}
#| label: fig-two
#| fig-cap: "A) PHQ-4 depression and anxiety scores against their respective ground-truth measures, the BDI-22 and the STAIT-5. Bayes factors in grey tell if there is a difference, for the same PHQ-4 score, between the original and the refined version (BFs < 1 suggest no difference and thus evidence for a comparability of the refined version with respect fo the original scale. Bayes factors in yellow represent how new in-between scores (0.5, 1.5, 2.5, ...) available with refined version differ from the adjacent scores (BFs > 3 suggest that half a point of difference on the refined PHQ-4 relates to a significant difference on the ground truth measure). BF < 1/3°, BF > 3\\*, BF > 10\\*\\*, BF > 30\\*\\*\\*. B) Bootstrapped distributions of the difference of correlation between the revised PHQ-4 scores and the original one for sub-clinical threshold scores of depression and anxiety. Positive differences suggest that the correlation between the ground-truth measure and the refined PHQ-4 score was stronger compared to the original version. C) Predictive power of the PHQ-4 scores on the presence of a depression or anxiety disorder. The upper plots show the relationship modelled by a logistic regression, while the above plots represent the ROC curves (in which a line further away from the diagonal represents a higher combination of sensitivity and specificity)."
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"

@@

### PHQ-4 Anxiety *vs.* STAIT-5

The linear regression predicting the STAIT-5 general score with the PHQ-4 anxiety score showed no interaction related to the PHQ-4 version ($\Delta Intercept_{\text{refined}} =  -0.02,~95\%~CI~[-0.15, 0.11],~t(832) =  -0.32,~p = 0.750$; $\Delta \beta_{\text{refined}} = 0.01,~95\%~CI~[-0.03, 0.05],~t(832) = 0.56,~p = 0.576$), suggesting no differences in the relationship pattern between the two versions.

Moreover, Bayesian *t*-tests comparing the STAIT-5 scores between the refined and the original version at each integer score yielded no evidence in favour of a significant difference. In other words, having the same score on the refined version as on the original version was related to the same outcome on the STAIT-5.

However, comparing in-between scores with adjacent scores yielded mixed results. Scoring 0.5 on the PHQ-4 anxiety was not significantly associated with a different level of STAIT-5 compared to scoring 0 (BF = 1.83), but was with scores of 1 (BF > 30). Similarly, there was no evidence that scoring 1.5 was different from scoring 1 (BF = 0.605), but strong evidence that it was different from scoring 2 (BF > 30).

@@

While the relationship pattern (i.e., the slope of the linear relationship) was not affected by the PHQ-4 version, we focused next on testing the difference in the strength (i.e., the precision) of the relationship, in particular at the lower end of the spectrum (i.e., for sub-clinical threshold scores of the BDI-II and STAIT-5). We bootstrapped (2000 iterations) the difference in correlation between the refined and the original version for each of the two ground-truth measures, separately for the BDI-II subsamples (minimal to mild <= 18; moderate to severe > 18) and the STAIT-5 subsamples (minimal to mild < 2; moderate to severe >= 2).

@@

For the STAIT-5, there was no difference in the correlation between the refined and the original version in the subclinical range of the STAIT-5. Surprisingly, we observed a stronger correlation between the refined PHQ-4 Anxiety score and the STAIT-5 in the moderate to severe range compared to the original version ($p_{one-sided} = 0.017$).
